---
title: "Bayesian Analysis of Beta-RD Statistic"
subtitle: "Rauf Salamzade"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

## Setup of R libraries 

```{r warning = FALSE, message = FALSE}
library(knitr)
library(rstan)
library(ggplot2)
library(corrplot)
library(RColorBrewer)
library(dplyr)
set.seed(12345)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Quick and Simple Preliminary Exploration of the Dataset

```{r}
data <- read.table('../../../../lsaBGC_Manuscript_and_Documentation/BetaRD_Bayesian_Shrinking/Staphylococcus_BetaRD.txt', header=T, sep='\t')

nb.cols <- 28
gcf.cols <- c(colorRampPalette(brewer.pal(9, "Set1"))(nb.cols), c("#000000"))
names(gcf.cols) <- c("betalactone", "butyrolactone", "CDPS", 
                     "cyclic-lactone-autoinducer", "ectoine","epipeptide", 
                     "hserlactone", "ladderane", "lanthipeptide-class-i", 
                     "lanthipeptide-class-ii", "lanthipeptide-class-iii", 
                     "lanthipeptide-class-iv", "lanthipeptide-class-v", "LAP", 
                     "linaridin", "NAPAA", "NRPS", "NRPS-like", "nucleoside", 
                     "phenazine", "RiPP-like", "RRE-containing", "sactipeptide", 
                     "siderophore", "T1PKS", "T3PKS", "terpene", "thiopeptide", 
                     "hybrid")

ggplot(data, aes(x=gcf_name, y=beta_rd)) + theme_classic() + 
  geom_boxplot(aes(fill=gcf_class),show.legend=F) + 
  scale_fill_manual(values=gcf.cols) + 
  geom_hline(yintercept=1.0, color='red', linetype='dashed', size=1.3) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("GCF") + ylab("Beta-RD") + ylim(0,2)

# downsample data using dplyr and select 1000 beta-rd from each gcf
downsampled_data <- data %>% group_by(gcf_id) %>% slice_sample(n=500)

ggplot(downsampled_data, aes(x=gcf_name, y=beta_rd)) + theme_classic() + 
  geom_boxplot(aes(fill=gcf_class),show.legend=F) + 
  scale_fill_manual(values=gcf.cols) + 
  geom_hline(yintercept=1.0, color='red', linetype='dashed', size=1.3) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("GCF") + ylab("Beta-RD") + ylim(0,2)
```
## Specificying the Model and Hierarchical Structuring

\begin{align*}
& \overline{\mu} \sim Norm(\mu_{0}, \sigma_{0}^2) \\
& \tau \sim half-t_{7}(1) \\
& \mu_{1} ... \mu_{J} \sim Norm(\overline{\mu}, \tau) \\
& \sigma \sim half-t_{7}(1) \\
& y_{j} \sim Norm(\mu_{j}, \sigma^{2}) \\
\end{align*}

## Setting Priors and Getting Input Ready for MCMC Analysis

```{r}
y <- downsampled_data$beta_rd
gcf_id <- downsampled_data$gcf_id

y_mean <- mean(y)
y_sd <- sd(y)
n <- length(y)
J <- length(unique(gcf_id))

# set parameters
mu0 <- 1.0
sigma0 <- y_sd
R1 <- 1
R2 <- 1
```

## Run MCMC Analysis

```{r}
#hier_model <- stan_model(file = "Hierarchical_MCMC.stan")
#save(hier_model, file='hier_model.Rdata')
load("../../../../lsaBGC_Manuscript_and_Documentation/BetaRD_Bayesian_Shrinking/hier_model.Rdata")
```

And here is the code to use the STAN model above:

```{r}
stan_input_data <- list(n = n, J = J, y = y, gcf_id = gcf_id, mu0 = mu0, 
                        sigma0 = sigma0, R1 = R1, R2 = R2)

#fit <- sampling(object = hier_model, data = stan_input_data, cpus=4, iter=2000,
#                pars = c("post_y"))
#save(fit, file='staphylococcus_analysis_fit.RData')
load("../../../../lsaBGC_Manuscript_and_Documentation/BetaRD_Bayesian_Shrinking/staphylococcus_analysis_fit.RData")

# Let's examine the Rhat values first (to see if the chains converged enough)
fit_summary <- summary(fit)
fit_summary_df <- as.data.frame(fit_summary$summary)

# print range of Rhats
print('Range of Rhats:')
print(range(fit_summary_df$Rhat))
# Uneccessary code to visualize histogram of Rhat
ggplot(fit_summary_df, aes(x=Rhat)) + geom_histogram(color='black', bins=10) + theme_bw()

```
We see Rhat values are well below 1.1, so we believe that the chains have converged! 

## Visualizing the Posterior Distributions

```{r}
post_y <- as.data.frame(extract(object = fit, pars = "post_y")[["post_y"]])
gcf_info_df <- distinct(data.frame(gcf_id = downsampled_data$gcf_id, 
                    gcf_name = downsampled_data$gcf_name, 
                    gcf_class = downsampled_data$gcf_class))
gcf_info_df <- gcf_info_df[order(gcf_id),]
gcf_names_ordered <- gcf_info_df$gcf_name
gcf_class_ordered <- gcf_info_df$gcf_class

posterior.data <- data.frame()
for (j in 1:J) {
  beta_rd_posterior <- as.vector(post_y[j][,1])
  gcf_name <- rep(gcf_names_ordered[j], length(beta_rd_posterior))  
  gcf_class <- rep(gcf_class_ordered[j], length(beta_rd_posterior))
  median_beta_rd <- rep(median(beta_rd_posterior), length(beta_rd_posterior))  
  posterior.row <- data.frame(beta_rd_posterior = beta_rd_posterior, 
                              gcf_name = gcf_name, 
                              gcf_class = gcf_class,
                              median_beta_rd = median_beta_rd)
  posterior.data <- rbind(posterior.data, posterior.row)
}

ggplot(posterior.data, aes(x=reorder(gcf_name, median_beta_rd), 
                           y=beta_rd_posterior, fill=gcf_class)) +
  theme_classic() + geom_boxplot(show.legend=F) + 
  scale_fill_manual(values=gcf.cols) + 
  geom_hline(yintercept=1.0, color='red', linetype='dashed', size=1.3) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("GCF") + ylab("Beta-RD") + ylim(0,2)

pdf("../../../../lsaBGC_Manuscript_and_Documentation/BetaRD_Bayesian_Shrinking/Staphylococcus_Posterior_BetaRD.pdf", height=5, width=9.38)
ggplot(posterior.data, aes(x=reorder(gcf_name, median_beta_rd), 
                           y=beta_rd_posterior, fill=gcf_class)) +
  theme_classic() + geom_boxplot(show.legend=F) + 
  scale_fill_manual(values=gcf.cols) + 
  geom_hline(yintercept=1.0, color='red', linetype='dashed', size=1.3) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  xlab("GCF") + ylab("Beta-RD") + ylim(0,2) + theme(text = element_text(size=20))
dev.off()
```
